---
title: "Data Understanding & Exploration - Self-Checkout Fraud Dataset"
author: "Philipp Altenbach, Ronny Grieder, Omar Rahiel, Emre Yelögrü"
format: html
editor: visual
execute:
  echo: true  # Show code in output by default
---

This document presents an initial exploratory analysis of the dataset related to self-checkout fraud detection. The focus is on understanding the structure of the `fraud.csv` dataset before proceeding with further analysis.

# Load and Inspection of the Data

```{r setup, include=FALSE}
# Load required libraries
library(data.table)  # Efficient data reading
library(dplyr)       # Data manipulation
library(ggplot2)     # Visualization
library(here)        # Using relative path names
```

```{r}
# Load dataset
data_path <- file.path(dirname(dirname(here())), "Data", "fraud.csv")
df <- fread(data_path)
```

```{r include=FALSE}
# Displaying the dimensions of the dataset
num_rows <- dim(df)[1]
num_cols <- dim(df)[2]

```

## Dimensions of the dataset

```{r}
# Dimensions of the dataset
cat("The dataset contains", num_rows, "rows and", num_cols, "columns.\n")

```

## Underlying data types

```{r}
# Display the data type of each attribute
column_info <- sapply(df, class)

# Print column names with their data types
for (col_name in names(column_info)) {
  print(paste("Column:", col_name, "- Data type:", column_info[col_name]))
}

```

## Summary of the whole dataset

```{r}
summary(df)
```

## Checking for missing values and duplicates

```{r include=FALSE}
# Count missing values per column
missing_values <- colSums(is.na(df))
# Count duplicate rows
duplicate_count <- nrow(df) - nrow(unique(df))

```

```{r}
cat("The dataset contains", sum(missing_values), "missing values and ", duplicate_count, " duplicates.\n")

```
